Squint: Lossy Hierarchical Compression on Symbol Streams

### 2012-12-11 tex/paper.tex
redid math section, improved some small sentences in results section

In order to make our algorithm for lossily compressing streams
as general as possible, we decided
to approach the problem of lossy compression by starting with a lossless
grammar inference algorithm. 
By building off of a grammar inference algorithm, we
hope to take advantage of some of the partial structures
that are visible in a CFG representation of a stream. 
In this project, we chose \Sequitur, but
any way of obtaining a grammar could be used. Then, we would transform
the grammar, introducing loss and perform compressions by simplifying
the grammar. Thus our compression system is a multi-step process:

@@

The previous algorithm updates the grammar sequentially, since each
variable replacement can have an effect on large parts of the grammar,
we will have to go through comparing all pairs again after the change.
An alternative algorithm would be to try to replace similar variables in
parallel. This would avoid many of the repeat pair-comparisons done in
\emph{Similarity}. In other words, the algorithm below puts 
similar variables
into equivalence class clusters, and then replaces each variable
in a cluster them with a representative
all at once each iteration.

\begin{algorithm}[h]
\caption{Cluster Lossifier Algorithm}
\label{cluster_alg}
\begin{algorithmic}[1]
\Procedure{Cluster-Sim}{$g$} \Comment{Simplify grammar g}
  \While{there are similar variables to replace}
    \State $clusters \gets ()$
    \For{$var_1 \rightarrow rhs_1 \in g$}
      \For{$c \in clusters$}
        \State $(var_2 \rightarrow rhs_2) \gets $ first variable in $c$
        \State $str_{i} \gets expand(rhs_i)$
        \If{$lsim(str_{1},str_{2})<\epsilon$}
          \State add $var_1$ to $c$
          \State break
        \EndIf
      \EndFor
      \If{$var_1$ hasn't been added to a cluster}
        \State Add $var_1$ to a new cluster in $clusters$
      \EndIf
    \EndFor
    \For{Cluster $c \in clusters$}
      \State replace all $var \in c$ with the smallest.
    \EndFor
  \EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\subsection{Runtime Estimates}
Given an input of length $n$, let $r$ be the number of rules generated
by Sequitur from the input. Let $R$ be the total number of variable
replacements we perform. For each replacement, we had to have searched though
up to all pairs of variables, and for each pair of variables expanding
them and calculating the levenshtein distance could take time linear in $n$.
Thus the runtime is $\boxed{O(R*r^2*n)}$.

From the \Sequitur paper \cite{sequitur} we
know that the number of rules in the grammar generated by Sequitur
is between $O(\sqrt{n})$ and $O(n)$. Thus the worst case runtime on the similarity
algorithm is $\boxed{O(n^4)}$. This is because we could have up to $r$ replacements.
However, in practice we will see that the runtime is much better
than $O(n^4)$. This is because we found $r$ for our test data to be slightly sublinear
even for complicated english text, and we found that we didn't quite make a linear
number of replacements $R$ with respect to $r$.
so our optimistic runtime could be anywhere 
from $O(n^4)$ to $O(n^{2})$ depending on how $r$ grows.

The inner loops of the Cluster are similar to those of the Similarity algorithm.
We iterate the entire clustering and replacement procedure $C$ times. Going through
each variable-cluster pair takes up to $O(r^2)$ time and calculating the levenshtein
distance takes $O(n)$ time. The time taken to actually perform the replacements each
iteration is absorbed into the $O(r^2)$ term. So, again we have a runtime of
$\boxed{O(C*r^2*n)}$. The big gain over the similarity algorithm is that $C$ turns out
to be nearly constant, since there are very few huge chains of reductions that are
only possible after other reductions have been performed, an in fact we are able to 
perform a very large number of the remaining reductions each time we perform an iteration
of the clustering algorithm. Thus since $C$ is nearly constant, the optimistic runtime
could be anywhere from $O(n^3)$ to $O(n^2)$.

CANT THINK OF PROOF OF NOLOOPS

@@

\subsection{Grammar Reductions}
What is their purpose? Why do we want them? What are they?

@@

\subsection{Similarity vs Clustering}
For most of our benchmark tests, we focused on two extreme kinds of data:
one with large amounts of repetition at all levels, and one with a word
and sentence level structure but with little obvious large-scale repetition.
One set of test data was constructed by repeating the same letter `a' n-times
but then adding some
texture to the data by changing 1/30 of the characters uniformly
at random to `*'. Call these test-cases \emph{rep-n}
Another set of test data was constructed from the first n characters of the
book of Genesis from the KJV. Call these test-cases \emph{gen-n}

@@

Below are the results on running Sequitur and Cluster, without
the grammar reductions, on the rep-n datasets. Time is the wall clock
running as measured on a 2.26 Ghz Intel Core 2 Duo Macbook Pro.
Ratio is the relative size of the grammar produced compared with
Sequitur, where the size of the grammar is the sum of the lengths of the
rhs's of the rules. Levdistance is the levenshtein distance between
the output of the lossified grammar and the original input string, it
is a kind of measure of the fidelity of our algorithms but doesn't
take into account how well patterns were preserved. For the rep-n tests,
proportion is the proportion of the final string produced that was
`*'. The closer to 1/30~0.033, the more accurate the lossifier was
in recreating the original texture of the input.

The columns of the data are prefixed by S or T depending on whether
they are measurements for sequitur or cluster respectively.

@@

Setting higher values for $\epsilon$ had diminishing returns for compression
for larger and larger file sizes, at the cost of drastically lowering the fidelity
of the output.

@@

Repeating the above tests on the \emph{gen-n} datasets yields the following:

Just as with the rep-n data, log-log regression tests yield the following
exponents for the runtime on the gen-n data.

@@

Again, the two algorithms have almost identical compression ratios and
their fidelity is about the same both subjectively and based on levenshtein
distance.

INCLUDE SAMPLE TEXT

@@

\subfloat[$\epsilon = .2$]{
\begin{tabular}{l|l|l|l|l||l|l|l|l}
Dataset & S-T(ime) & S-R(atio) & S-L(evdistance) & S-P(roportion) & C-T & C-R & C-L & C-P \\
\hline
rep-100   & 0.002 & 30/34    & 0.040 & 0.05 & 0.002 & 30/34    & 0.040 & 0.05\\
rep-400   & 0.019 & 60/76    & 0.035 & 0.02 & 0.009 & 60/76    & 0.035 & 0.02\\
rep-1600  & 0.350 & 123/194  & 0.355 & 0.03 & 0.055 & 123/194  & 0.206 & 0.01\\
rep-6400  & 3.942 & 280/448  & 0.387 & 0.08 & 0.295 & 274/448  & 0.543 & 0.09\\
rep-12800 & 42.21 & 638/1064 & 0.491 & 0.03 & 1.105 & 632/1064 & 0.457 & 0.01\\
rep-25600 & 124.5 & 922/1482 & 0.565 & 0.12 & 2.597 & 926/1482 & 0.437 & 0.007\\
\end{tabular}
}
\caption{rep-n Tests}
\label{seqclus_r_t}
\end{table}

@@

\subfloat[$\epsilon = .2$]{
\begin{tabular}{l|l|l|l||l|l|l}
Dataset & S-T(ime) & S-R(atio) & S-L(evdistance) & C-T & C-R & C-L \\
\hline
gen-100   & 0.004 & 89/89    & 0.00 &0.002 & 89/89    & 0.00 \\
gen-400   & 0.085 & 245/258  & 0.03 &0.053 & 245/258  & 0.03 \\
gen-1600  & 1.292 & 664/697  & 0.03 &0.874 & 664/697  & 0.03 \\
gen-6400  & 73.83 & 2173/2348& 0.05 &14.02 & 2171/2348& 0.05\\
gen-12800 & 549.2 & 4197/4253& 0.05 &45.64 & 4186/4253& 0.05\\
gen-25600 & 3939  & 7405/8060& 0.06 &224.9 & 7374/8060& 0.06\\
\end{tabular}
}
\caption{gen-n Tests}
\label{seqclus_g_t}
\end{table}

@@

1:26 And God said, Let us make man in our image, after our likeness:
and let them have dominion over the fish of the sea, and over the fowl
of the air, and over the cattle, and over all the earth, and over
every creeping thing that creepeth upon the earth.
\end{lstlisting}

>>>>

In order to make our algorithm for lossily compressing streams
as general as possible, we decided
to approach the problem of lossy compression by starting with a lossless
grammar inference algorithm. 
By building off of a grammar inference algorithm, we
hope to take advantage of some of the partial structures
that are visible in a CFG representation of a stream. 
In this project, we chose \Sequitur, but
any way of obtaining a grammar could be used. Then, transformed
the grammar, introducing loss and perform compressions by simplifying
the grammar. Thus our compression system is a multi-step process:

@@

The previous algorithm updates the grammar sequentially, since each
variable replacement can have an effect on large parts of the grammar,
we will have to go through comparing all pairs again after the change.
An alternative algorithm would be to try to replace similar variables in
parallel. This would avoid many of the repeat pair-comparisons done in
\emph{Similarity}. In other words, the algorithm below puts 
similar variables
into equivalence class clusters, and then replaces each variable
in a cluster them with the smallest representative
all at once each iteration.

\begin{algorithm}[h]
\caption{Cluster Lossifier Algorithm}
\label{cluster_alg}
\begin{algorithmic}[1]
\Procedure{Cluster-Sim}{$g$} \Comment{Simplify grammar g}
  \While{there are similar variables to replace}
    \State $clusters \gets ()$
    \For{$var_1 \rightarrow rhs_1 \in g$}
      \For{$c \in clusters$}
        \State $(var_2 \rightarrow rhs_2) \gets $ first variable in $c$
        \State $str_{i} \gets expand(rhs_i)$
        \If{$lsim(str_{1},str_{2})<\epsilon$}
          \State add $var_1$ to $c$
          \State break
        \EndIf
      \EndFor
      \If{$var_1$ hasn't been added to a cluster}
        \State Add $var_1$ to a new cluster in $clusters$
      \EndIf
    \EndFor
    \State save a mapping $m$ from variables to sizes
    \For{Cluster $c \in clusters$}
      \State replace all $var \in c$ with the smallest (smallest w.r.t. $m$).
    \EndFor
  \EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\subsection{Proof of Grammar Consistency}

One risk in replacing variables in a CFG is that one could introduce
cyclical expansion in the CFG, such as if 
$A\rightarrow aB$ and $B\rightarrow bA$. In this 
section we will argue that neither of our algorithms do so if the
original grammar was well formed.

We can view the CFG as a DAG where for variables $V_1$, $V_2$, we add an edge
whenever $V_1$ contains $V_2$ in its rhs. When sizes are compuated
on a CFG before we modify it, if $V_1 \rightarrow V_2$ then $size(V_1) > size(V_2)$.
Replacing a variable $V_1$ with
$V_2$ has the same effect as redirecting all in-edges to $V_1$ onto $V_2$.

Suppose for the sake of contradiction that the Similarity algorithm
created a cycle at one step by redirecting an edge originally pointing 
at $A$ to $B$. This would mean that there was a path going from 
$B$ to $A$. However, by comparing the sizes of $A$ and $B$ we know that
$A$ must have been bigger than $B$. Thus there cannot have been a path from $B$
to $A$, and we have a contradiction.

For the Cluster algorithm the analysis is a bit more involved, since
to avoid recomputation we only calculate the sizes of the variables once
before doing all of the replacements. First note that even as we modify
the DAG, all of the edges we create in the DAG will respect the ordering of
sizes that we calculated before we did any replacements. This is because
each time we redirect an edge from $X\rightarrow A$ to $X\rightarrow B$
$size(X)>size(A)>size(B)$ and transitivity will preserve the ordering.

Suppose then that we created a cycle during Cluster
by redirecting an edge from $X \rightarrow A$ to $X \rightarrow B$.
This means that there must have been a path from $B$ to $X$. However,
even in its intermediate state we know that the DAG respects the
$size$ ordering, and $size(X) > size(A) > size(B)$, so there cannot
be a path from $B$ to $X$. This is a contradiction and we are done.

\subsection{Runtime Analysis}
Given an input of length $n$, let $r$ be the number of rules generated
by Sequitur from the input. 
In the analysis of the Similarity algorithm, let $R$ be the total number of variable
replacements we perform. For each replacement, we had to have searched though
up to all pairs of variables, and for each pair of variables expanding
them and calculating the levenshtein distance could take time linear in $n$.
Performing the replacement only takes time linear in $r$.
Thus the runtime is $\boxed{O(R*r^2*n)}$.

From the \Sequitur paper \cite{sequitur} we
know that the number of rules in the grammar generated by Sequitur
is between $O(1)$ and $O(n)$. Thus the worst case runtime on the similarity
algorithm is $\boxed{O(n^4)}$. This is because we could have up to $r$ replacements.
However, in practice we will see that the runtime is much better
than $O(n^4)$. There are a variety of possible reasons for this:
in our test data, $r$ didn't grow quite linearly in $n$, $R$ could grow much
slower than $r$, and the use of native code in calculating the levenshtein distance 
made the its contribution much less noticeable for the input sizes we tested.

The inner loops of the Cluster are similar to those of the Similarity algorithm.
We iterate the entire clustering and replacement procedure $C$ times. Going through
each variable-cluster pair takes up to $O(r^2)$ time and calculating the levenshtein
distance takes $O(n)$ time. The time taken to actually perform the replacements each
iteration is absorbed into the $O(r^2)$ term. So, again we have a runtime of
$\boxed{O(C*r^2*n)}$. Just as with the similarity algorithm,
a variety of factors make the runtime much better than the
$\boxed{O(n^4)}$ worst case scenario. 
The big gain over the similarity algorithm is that $C$ turns out
to be nearly constant, since there are very few huge chains of reductions that are
only possible after other reductions have been performed, an in fact we are able to 
perform a very large number of the remaining reductions each time we perform an iteration
of the clustering algorithm. Thus a more reasonable upper bound for the runtime
of cluster would be $\boxed{O(n^3)}$

@@

\subsection{Grammar Reductions}
What is their purpose? Why do we want them? What are they?
Grammar reductions are a way of taking a CFG, generated possibly through
ad-hoc techniques, and systematically removing internal redundancies
like duplicate bigrams in Sequitur. They make the CFG representation smaller
but don't introduce any loss. We want them since our lossifier algorithms
often make our grammars redundant, even though the each step we take
in the lossifier makes the grammars smaller.

@@

\subsection{Similarity vs Clustering}
For most of our benchmark tests, we focused on two extreme kinds of data:
one with large amounts of repetition at all levels, and one with a word
and sentence level structure but with little obvious large-scale repetition.
One set of test data was constructed by repeating the same letter `a' n-times
but then adding some
texture to the data by changing 1/30 of the characters uniformly
at random to `*'. Call these test-cases \emph{rep-n}
Another set of test data was constructed from the first n characters of the
book of Genesis from the KJV version of the Bible. 
Call these test-cases \emph{gen-n}

@@

In table \ref{simclus_r_t} we report the results on running Sequitur 
and Cluster, without
the grammar reductions, on the rep-n datasets. Time is the wall clock
running as measured on a 2.26 Ghz Intel Core 2 Duo Macbook Pro.
Ratio is the relative size of the grammar produced compared with
Sequitur, where the size of the grammar is the sum of the lengths of the
rhs's of the rules. Levdistance is the levenshtein distance between
the output of the lossified grammar and the original input string, it
is a kind of measure of the fidelity of our algorithms but doesn't
take into account how well patterns were preserved. For the rep-n tests,
proportion is the proportion of the final string produced that was
`*'. The closer to 1/30~0.033, the more accurate the lossifier was
in recreating the original texture of the input.

The columns of the data are prefixed by S or T depending on whether
they are measurements for Similarity or Cluster respectively.

@@

The actual compression ratio was respectable, with both
Similarity and Cluster coming in at nearly half the size of the Sequitur CFG.
Setting higher values for $\epsilon$ had diminishing returns for compression
for larger and larger file sizes, at the cost of drastically lowering the fidelity
of the output.

@@

Repeating the above tests on the \emph{gen-n} datasets yields the results in table
\ref{simclus_g_t}

Just as with the rep-n data, log-log regression tests yield the following
exponents for the runtime on the gen-n data, again with very high values of
$R^2$.

@@

Just as in the rep-n results, the two algorithms have 
almost identical compression ratios and
their fidelity is about the same both subjectively and based on levenshtein
distance. For an example of the text generated on one of the gen-n datasets,
see the appendix.

@@

\subfloat[$\epsilon = .2$]{
\begin{tabular}{l|l|l|l|l||l|l|l|l}
Dataset & S-T(ime) & S-R(atio) & S-L(evdistance) & S-P(roportion) & C-T & C-R & C-L & C-P \\
\hline
rep-100   & 0.002 & 30/34    & 0.040 & 0.05 & 0.002 & 30/34    & 0.040 & 0.05\\
rep-400   & 0.019 & 60/76    & 0.035 & 0.02 & 0.009 & 60/76    & 0.035 & 0.02\\
rep-1600  & 0.350 & 123/194  & 0.355 & 0.03 & 0.055 & 123/194  & 0.206 & 0.01\\
rep-6400  & 3.942 & 280/448  & 0.387 & 0.08 & 0.295 & 274/448  & 0.543 & 0.09\\
rep-12800 & 42.21 & 638/1064 & 0.491 & 0.03 & 1.105 & 632/1064 & 0.457 & 0.01\\
rep-25600 & 124.5 & 922/1482 & 0.565 & 0.12 & 2.597 & 926/1482 & 0.437 & 0.007\\
\end{tabular}
}
\caption{rep-n Tests}
\label{simclus_r_t}
\end{table}

@@

\subfloat[$\epsilon = .2$]{
\begin{tabular}{l|l|l|l||l|l|l}
Dataset & S-T(ime) & S-R(atio) & S-L(evdistance) & C-T & C-R & C-L \\
\hline
gen-100   & 0.004 & 89/89    & 0.00 &0.002 & 89/89    & 0.00 \\
gen-400   & 0.085 & 245/258  & 0.03 &0.053 & 245/258  & 0.03 \\
gen-1600  & 1.292 & 664/697  & 0.03 &0.874 & 664/697  & 0.03 \\
gen-6400  & 73.83 & 2173/2348& 0.05 &14.02 & 2171/2348& 0.05\\
gen-12800 & 549.2 & 4197/4253& 0.05 &45.64 & 4186/4253& 0.05\\
gen-25600 & 3939  & 7405/8060& 0.06 &224.9 & 7374/8060& 0.06\\
\end{tabular}
}
\caption{gen-n Tests}
\label{simclus_g_t}
\end{table}

@@

1:26 And God said, Let us make man in our image, after our likeness:
and let them have dominion over the fish of the sea, and over the fowl
of the air, and over the cattle, and over all the earth, and over
every creeping thing that creepeth upon the earth.
